---
title: "RFM Analysis"
author: "Md Shamsul Hoque Khan"
date: "2023-08-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to RFM Analysis

RFM (Recency, Frequency, Monetary) analysis is a customer segmentation technique that helps marketers understand the behavior of their customers. Each customer is scored on three dimensions:

Recency: How recently the customer made a purchase
Frequency: How often the customer makes a purchase
Monetary: How much the customer spends
Through clustering these metrics, businesses can identify customer groups and design targeted marketing strategies.

# Preprocessing
## Load Required Libraries

```{r}
# Load tidyverse for data manipulation and ggplot2 for plotting
library(tidyverse)

# Load readxl for reading Excel files
library(readxl)

```

## Load the Data

```{r}
# Load the sample_superstore data set downloaded from Tableau Public
suppressWarnings({
  df <- read_excel("sample_superstore.xls")
})


```

# Data Exploration

## Check for Missing Values

```{r}
# Check for any missing values in the data
sum(is.na(df))

```

No missing values are found in the dataset.


## Check Data Structure

```{r}
# Examine the structure of the dataset
str(df)

```

The data columns are in the correct format.

# RFM Calculation

## Recency Calculation

```{r}
# Identify the latest date in the dataset
max(df$`Order Date`)

```


The dataset contains data up to December 2017.

```{r}
# Calculate recency for each customer
reference_date <- as.Date(max(df$"Order Date")) + 1

recency_df <- df %>%
  group_by(`Customer ID`) %>%
  summarise(Recency = as.numeric(difftime(reference_date, max(`Order Date`), units="days"))) %>%
  arrange(Recency)

tail(recency_df)

```


## Frequency Calculation

```{r}
# Calculate the frequency of purchases for each customer
frequency_df <- df %>%
  group_by(`Customer ID`) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency))

```

## Monetary Value Calculation

```{r}
# Calculate the monetary value (based on Sales) for each customer
monetary_df <- df %>%
  group_by(`Customer ID`) %>%
  summarise(Monetary = sum(Sales))

```

## Combine RFM Metrics

```{r}
# Combine Recency, Frequency, and Monetary metrics into one RFM DataFrame
rfm_df <- recency_df %>%
  inner_join(frequency_df, by = "Customer ID") %>%
  inner_join(monetary_df, by = "Customer ID")

# Standardize the metrics for further clustering
scaled_rfm <- as.data.frame(scale(rfm_df[, c('Recency', 'Frequency', 'Monetary')]))
scaled_rfm_with_ID <- data.frame("Customer ID" = rfm_df$`Customer ID`, scaled_rfm)

head(scaled_rfm_with_ID)

```


# Clustering

## Determine Number of Clusters

```{r}
# Use the elbow method to identify the optimal number of clusters
set.seed(12)
wss <- numeric()

for (i in 1:10){
  kmeans_model <- kmeans(scaled_rfm_with_ID[,-1], centers = i, nstart = 20)
  wss[i] <- kmeans_model$tot.withinss
}

```


## Elbow Plot

```{r}
# Plot the elbow graph to determine the optimal number of clusters
ggplot(data.frame(Clusters = 1:10, WSS = wss), aes(x = Clusters, y = WSS)) +
  geom_point() +
  geom_line() +
  ggtitle("Elbow Method to Determine Optimal Number of Clusters") +
  xlab("Number of Clusters") +
  ylab("Within-Cluster Sum of Squares") +
  scale_x_continuous(breaks = seq(1, 10, by = 1))

```

Based on the elbow plot, 3 clusters appear to be optimal.


## K-Means Clustering

```{r}
# Run K-means clustering algorithm with the optimal number of clusters (3)
set.seed(12)
kmeans_model <- kmeans(scaled_rfm_with_ID[,-1], centers = 3, nstart = 20)

# Add the cluster labels to the original RFM DataFrame
rfm_df[, "Cluster"] <- kmeans_model$cluster

```


# Cluster Analysis

## Cluster Descriptions

```{r}
# Calculate the average Recency, Frequency, and Monetary value for each cluster
group_rfm_df <- rfm_df %>% 
  group_by(Cluster) %>% 
  summarise(
    Recency = mean(Recency),
    Frequency = mean(Frequency),
    Monetary = mean(Monetary)
  )

group_rfm_df

```


## Label Customer Types

```{r}
# Label customer types based on their cluster
rfm_df_labeled <- rfm_df %>% 
  mutate(
    Customer_type = case_when(Cluster == 1 ~ "Great Customer",
                              Cluster == 2 ~ "Disloyal Customer",
                              TRUE ~ "Average Customer")
  )

head(filter(rfm_df_labeled, Cluster == 2))

```

## Plotting Customer Types

```{r}
# Bar plot to show the distribution of customer types
ggplot(rfm_df_labeled, aes(x = Customer_type, fill = Customer_type)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..),
    stat = 'count',
    position = position_stack(vjust = 0.5)
  ) +
  ggtitle("Distribution of Customer Types") +
  xlab("Customer Type") +
  ylab("Frequency") +
  scale_fill_manual(values = c("Great Customer" = "dark green", "Average Customer" = "blue", "Disloyal Customer" = "red"))

```

# Conclusion

Based on the RFM analysis, we observe a large number of Average Customers and a reasonable number of Great Customers. Only about 14.75% of the customer base is classified as disloyal. The store should focus its marketing efforts on retaining Great Customers and converting Average Customers to Great Customers.










